---
title: "Can simple, accessible measures<br>reliably predict body fat %?"
subtitle: "L07G01 - Modelling Body Fat"
author: "Lara Burns, Karin Gao, Qiuyu Liu, Bryan Chan and Hanna Boudara"
format: 
  revealjs:
    auto-slide: 20000
    embed-resources: true
    transition: slide
    css: "quarto presentation styles.css"
---

```{r, message=FALSE}
library(tidyverse)
library(readr)
# install.packages("modelsummary")
library(modelsummary)
library(ggcorrplot)
# install.packages("caTools")
library(caTools)
library(car)
# install.packages("sjPlot")
library(sjPlot)
library(insight)
# install.packages("lmSubsets")
library(lmSubsets)
library(broom)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
theme_set(theme_bw())
```

## Why Predict Body Fat %?

::: {style="font-size:1.7rem; line-height:1.8"}
- Accurate health metric 
- Helps manage risks (cardiovascular disease, diabetes) 
- Informs targeted fitness decisions 
:::

::: columns 
::: {.column width="38%"}
::: {.callout-tip title="Goal"}
Use simple body measurements:

- Age
- Height
- Weight
- Circumferences

:::
:::

::: {.column width="62%"}
::: {.callout-warning title="However, people usually rely on:"}
- **Oversimplified indicators** (BMI)
- **Expensive / inaccessible** tests (DEXA, hydrostatic)
:::

:::
:::

## Data Cleaning and Splitting

The original data had no missing values. For the variable `Weight`, which was in pounds, was converted to kilograms and `Height`, which was in inches, to centimeters. After this data cleaning, we had the following data set to build our models upon:

```{r}
# Read in data and inspect
bodyfat <- read_tsv("bodyfat.txt") 
```

```{r}
bodyfat$Age <- as.integer(bodyfat$Age)
bodyfat$Weight <- bodyfat$Weight/2.205
bodyfat$Height <- bodyfat$Height*2.54
glimpse(bodyfat)

```

Our models were built on 80% of the data set, designated as the training data, and the models' performance was built on the remaining 20% of the data, designated as the test data.

```{r}
set.seed(71) # Set random seed

split <- sample.split(bodyfat$Pct.BF, SplitRatio = 0.8)
train <- subset(bodyfat, split == TRUE)
test <- subset(bodyfat, split == FALSE)
```

## Exploratory Data Analysis

```{r}
# install.packages("patchwork")
library(patchwork)

# Histogram of Body Fat %
p1 <- ggplot(bodyfat, aes(x = Pct.BF)) +
  geom_histogram(binwidth = 3, boundary = 0,
                 fill = "lightblue", color = "white", alpha = 0.9) +
  labs(title = "Distribution of Body Fat Percentage",
       x = "Body Fat (%)", y = "Count") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5))

# Scatter plots
vars_to_plot <- c("Density", "Abdomen","Waist","Chest","Hip","Weight","Thigh",
                  "Neck","Knee","Bicep","Forearm","Wrist","Age","Ankle","Height")

p2 <- bodyfat %>%
  select(Pct.BF, all_of(vars_to_plot)) %>%
  pivot_longer(cols = all_of(vars_to_plot),
               names_to = "variable", values_to = "value") %>%
  drop_na() %>%
  ggplot(aes(x = value, y = Pct.BF)) +
  geom_point(alpha = 0.5, colour = "lightblue", size = 0.9) +
  geom_smooth(method = "lm", se = TRUE, colour = "red", linewidth = 0.7) +
  facet_wrap(~ variable, scales = "free_x", ncol = 5) +
  labs(x = NULL, y = "Body Fat (%)",
       title = "Relationships between Body Fat % and Predictors") +
  theme_classic() + 
  theme(legend.position = "top",
        strip.text = element_text(size = 10),
        plot.title = element_text(size = 14, hjust=0.5))

# Combining graphs together
p1 + p2 + plot_layout(widths = c(1, 2))
```

## Full Linear Model

```{r}
M1 = lm(Pct.BF ~.-Density, data=train) # Full Model without density
round(summary(M1)$coef, 3)
```

Our resulting model is:

$$\widehat{\text{Pct.BF}} = 484.598 + 0.008(\text{Age})+ 0.016(\text{Weight})- 0.021(\text{Height}) \\+ 0.027(\text{Neck})- 0.003(\text{Chest})- 0.006(\text{Abdomen})+ 0.024(\text{Hip})- 0.001(\text{Thigh}) \\- 0.048(\text{Knee})+ 0.016(\text{Ankle})- 0.003(\text{Bicep})+ 0.005(\text{Forearm})- 0.098(\text{Wrist})$$

```{r}
# Helper function
model_footer <- function(model) {
  s <- summary(model)
  
  # Extract p-value from F-statistic
  p_val <- pf(s$fstatistic[1], s$fstatistic[2], s$fstatistic[3], lower.tail = FALSE)
  
  # Print formatted summary
  cat(
    sprintf("Residual standard error: %.3f on %d degrees of freedom\n", s$sigma, s$df[2]),
    sprintf("Multiple R-squared: %.4f, Adjusted R-squared: %.4f\n", s$r.squared, s$adj.r.squared),
    sprintf("F-statistic: %.1f on %d and %d DF, p-value: %.3g\n",
            s$fstatistic["value"], s$fstatistic["numdf"], s$fstatistic["dendf"], p_val)
  )
}
```

```{r}
model_footer(M1)
```

## Full Linear Model (Continued)

In the full model, majority of predictors are statistically insignificant (with `Pr(>|t|)` \> 0.05) and this is potentially due to multi-collinearity between the predictors.

```{r}
alias(M1)
```

This shows that waist and abdomen are nearly perfectly correlated. This would make model predictions unstable, hence we can try different models selection methods.

## Model with Interaction Term

```{r}
lm_bh = lm(Pct.BF ~ Abdomen + Waist + Chest + Hip + Abdomen:Chest, data = train)
summary(lm_bh)
```

## Model with Interaction Term (Continued)

```{r}
aug_bh = augment(lm_bh)

p1 <- ggplot(aug_bh, aes(.fitted, .std.resid)) +
geom_point(alpha = 0.6) +
geom_hline(yintercept = 0, color = "red") + 
labs(x = "Fitted values", y = "Standardized Residuals", title = "Residuals vs Fitted for Abdomen x Chest") +
geom_smooth(method = "loess", se = FALSE)

p2 <- ggplot(aug_bh, aes(sample = .std.resid)) +
stat_qq() +
stat_qq_line() +
labs(title = "Q-Q Plot of Residuals")
p1 + p2
```

```{r}
pred_bh = predict(lm_bh, newdata = test)
RMSE = function(actual, pred) sqrt(mean((actual - pred)^2))
rmse_bh = RMSE(test$Pct.BF, pred_bh)
rmse_bh

```

```{r}
lm_bh_noint = lm(Pct.BF ~ Abdomen + Waist + Chest + Hip, data = train)
anova(lm_bh_noint, lm_bh)  # Likelihood ratio test
AIC(lm_bh_noint, lm_bh)    # Compare the model selection criteria
```

```{r}
# Refit without interaction, bc high p-value
lm_bh_final <- lm(Pct.BF ~ Abdomen + Waist + Chest + Hip, data = train)

# RMSE
pred_bh_final <- predict(lm_bh_final, newdata = test)
rmse_bh_final <- RMSE(test$Pct.BF, pred_bh_final)
rmse_bh_final
```

## Multiple Body Sections Model

::::: columns
::: column
This model was developed to examine whether a holistic view across different areas of the body can help to produce an accurate prediction of body fat.

```{r}
correlations <- cor(bodyfat[, c("Pct.BF", "Abdomen", "Chest", "Waist", "Hip",
                                "Bicep", "Forearm", "Wrist",
                                "Neck", 
                                "Thigh", "Knee", "Ankle")], use = "complete.obs")

correlations["Pct.BF", ]
```

The variables most correlated with body fat percentage from each body section are: `Abdomen` (torso), `Bicep` (arms), `Neck` (neck/head), and `Thigh` (legs). The resulting model using these:

```{r}
lm1 <- lm(Pct.BF ~ Abdomen + Bicep + Neck + Thigh, data = train)
round(summary(lm1)$coef, 2)
```

$\widehat{\text{Pct.BF}} = -18.10 + 0.85 \, \text{Abdomen} + 0.02 \, \text{Bicep} - 0.96 \, \text{Neck} - 0.10 \, \text{Thigh}$
:::

::: column
```{r}
aug_mbp = augment(lm1)

p1 <- ggplot(aug_mbp, aes(.fitted, .std.resid)) +
geom_point(alpha = 0.6) +
geom_hline(yintercept = 0, color = "red") + 
labs(x = "Fitted values", y = "Standardized Residuals", title = "Residuals vs Fitted for Abdomen x Chest") +
geom_smooth(method = "loess", se = FALSE)
p1
```

```{r}
p2 <- ggplot(aug_mbp, aes(sample = .std.resid)) +
stat_qq() +
stat_qq_line() +
labs(title = "Q-Q Plot of Residuals")
p2
```
:::
:::::

## Backwards Selection

```{r}
step.back.aic = step(M1, 
					direction = "backward", 
					trace = FALSE) 
round(summary(step.back.aic)$coef, 3)
```

Our resulting model is:

$$\widehat{Pct.BF} = 9.551 - 0.114(Height) - 0.625(Neck) + 0.940(Abdomen) - 1.217(Wrist)$$

```{r}
model_footer(step.back.aic)
```

## Forward Selection

```{r}
M0 = lm(Pct.BF ~ 1, data=train) # Null model
step.fwd.aic = step(M0, scope = list(lower = M0, upper = M1),
                    direction = "forward", trace = FALSE)
round(summary(step.fwd.aic)$coef, 3)

```

```{r}
fitted = fitted(step.fwd.aic)
res = residuals(step.fwd.aic)

df_forward = data.frame(Fitted = fitted, Residuals = res)


p1 = ggplot(data = df_forward, aes(x = Fitted, y = Residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype="dashed") +
  labs(x = "Fitted values", y = "Residuals") +
  geom_smooth(method = "loess", se = FALSE)

p2 = ggplot(df_forward, aes(sample = Residuals)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q Plot for Forward Model", x = "Theoretical Quantities", y = "Sample Quantities")

p1+p2

```

Our resulting model is:

$$
\widehat{\text{Pct.BF}} = 486.473 - 439.635(\text{Density}) - 0.025(\text{Height}) + 0.014(\text{Weight})
$$

## Comparing Stepwise Models

```{r}
sjPlot::tab_model(step.fwd.aic, step.back.aic, 
                  show.ci = FALSE, 
                  show.aic = TRUE,
                  dv.labels = c("Forward model", "Backward model"))
```

## Exhaustive Subset Search (BIC)

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.height=5}
exh <- lmSubsets(Pct.BF ~ . - Density, data = train, nbest = 15)
plot(exh)

best_bic <- lmSelect(exh, criterion = "BIC")
sel_coefs <- coef(best_bic)                 
selected_vars <- setdiff(names(sel_coefs), "(Intercept)")
k <- length(selected_vars)

## Refit using exactly those predictors
best_model <- lm(reformulate(selected_vars, response = "Pct.BF"), data = train)

# quick metrics
adjR2_train <- summary(best_model)$adj.r.squared
pred_test   <- predict(best_model, newdata = test)
rmse_test   <- sqrt(mean((test$Pct.BF - pred_test)^2))
```
::: {.callout-tip title="Takeaway"}
**BIC minimum at `r k` predictors** → compact model.

Selected: `r paste(selected_vars, collapse = ", ")`.
:::

## Selected Exhaustive Model (BIC)
::: columns
::: {.column width="42%"}
```{r, echo=FALSE, message=FALSE, warning=FALSE}


tab <- tidy(best_model) |>
  filter(term != "(Intercept)") |>
  transmute(Predictor = term,
            Estimate  = round(estimate, 2),
            p         = sprintf("%.3f", p.value))

kable(tab, align = "lrr") |>
  kable_styling(full_width = FALSE, position = "left", font_size = 18)
```
:::

::: {.column width="58%"} 
```{r}
aug <- augment(best_model)
ggplot(aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = "Residuals vs Fitted",
       x = "Fitted values", y = "Residuals") +
  theme_classic(base_size = 12)
```
:::
:::
```{r}
coefs <- coef(best_model)
```

::: {.callout-note title="Summary"}
**Resultant Model (BIC)**: $\widehat{\text{Pct.BF}} = -30.32 - 0.23\cdot\text{Weight} +2.47\cdot\text{Waist}-1.2\cdot\text{Wrist}$

- ***Residuals***: Slight S-shape but no severe pattern; spread roughly constant.
- ***Train adj*** $R^2$: `r sprintf('%.2f', adjR2_train)`.
- ***Test RMSE***: `r sprintf('%.2f', rmse_test)`.
:::

## Comparing All Models

To find the best overall model, all developed models were tested against the remaining 20% of the data set aside. This allowed us to test the models against unseen data.

```{r}
y_test <- test$Pct.BF

# Predictions on test set
pred_full <- predict(M1, newdata = test)
pred_fwd <- predict(step.fwd.aic, newdata = test)
pred_back <- predict(step.back.aic, newdata = test)
pred_exh <- predict(best_model, newdata = test)
pred_mbp <- predict(lm1, newdata = test)
pred_bp <- predict(lm_bh, newdata = test)

# Number of predictors for each model
p_full <- length(coef(M1)) - 1
p_fwd  <- length(coef(step.fwd.aic)) - 1
p_back <- length(coef(step.back.aic)) - 1
p_exh  <- length(coef(best_model)) - 1
p_mbp  <- length(coef(lm1)) - 1
p_bp <- length(coef(lm_bh)) - 1

# Helper function
RMSE <- function(actual, predicted) sqrt(mean((actual - predicted)^2))
MAE <- function(actual, predicted) mean(abs(actual - predicted))
adj_R2 <- function(actual, predicted, p) {
  n <- length(actual)
  r2 <- cor(actual, predicted)^2
  1 - (1 - r2) * (n - 1) / (n - p - 1)
}

results <- data.frame(
Model = c("Full Model", "Model with Interactive Term", "Multiple Body Parts", "Forward Stepwise",
"Backward Stepwise", "Exhaustive (BIC)"),
RMSE = c(RMSE(y_test, pred_full),
RMSE(y_test, pred_bp),
RMSE(y_test, pred_mbp),
RMSE(y_test, pred_fwd),
RMSE(y_test, pred_back),
RMSE(y_test, pred_exh)),
MAE = c(MAE(y_test, pred_full),
MAE(y_test, pred_bp),
MAE(y_test, pred_mbp),
MAE(y_test, pred_fwd),
MAE(y_test, pred_back),
MAE(y_test, pred_exh)),
R2 = c(adj_R2(y_test, pred_full, p_full),
adj_R2(y_test, pred_bp, p_bp),
adj_R2(y_test, pred_mbp, p_mbp),
adj_R2(y_test, pred_fwd, p_fwd),
adj_R2(y_test, pred_back, p_back),
adj_R2(y_test, pred_exh, p_exh))
)

library(gt)

results |>
  rename(`Adj. R^2` = R2) |>
  gt() |>
  tab_header(
    title = "Model Performance Comparison"
  ) |>
  fmt_number(columns = c(RMSE, MAE, `Adj. R^2`), decimals = 2)
```

It can be seen that the Forward Stepwise model performed the most accurately against this test data, with the lowest Residual Mean Squared of $4.17$ as well as the highest adjusted $R^2$ value of $0.7$

This means that this model predicts body fat percentage at an average of

$$
\sqrt(4.17) = 2.04\% 
$$

off the actual data points, which is fairly accurate.

## Best Overall Model

::::: columns
::: column
Revisiting the Forward Stepwise model:

```{r}
round(summary(step.fwd.aic)$coef, 2)
```

$$
\widehat{\text{Pct.BF}} = -32.68 + 2.59(\text{Weight}) - 0.98(\text{Wrist}) \\ + 0.25(\text{Bicep}) - 0.51(\text{Neck}) + 0.33(\text{Forearm})
$$

-   **Waist Coefficient: 2.59**

    As waist circumference increases by 1 cm, body fat increases by \~2.6%, holding other variables constant. This is the largest coefficient, suggesting Waist is the strongest predictor in predicting body fat percentage. It is also highly significant, with a p-value of 0.00.
:::

::: column
-   **Weight, Wrist and Neck Coefficients**

    An increase in the circumference of these factors all contribute to a slight decrease in body fat percentage. Counter-intuitive, but this could reflect collinearity with other predictors like waist.

-   **Bicep and Forearm Coefficients**

    A 1cm increase in bicep/forearm circumference is associated with a slight increase in body fat percentage.

Waist and weight can be seen to be statistically significant (\< 0.05). Wrist and Neck are marginally significant (0.05–0.10), which could be significant at lower confidence levels. On the other hand, bicep and forearm are not significant with a very high p-value (\> 0.10).
:::
:::::

## Assumptions of the Model

## Graphing of Residuals + QQ Plot (To be used for best overall model)

```{r}
# install.packages("broom")
library(broom)

aug <- augment(lm1)

p1 <- ggplot(aug, aes(.fitted, .std.resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "blue") +
  labs(x = "Fitted", y = "Std. Residual")

p2 <- ggplot(aug, aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line()

# Combine plots side by side
p1 + p2
```

## Limitations

-   Include recommendations for future research

## Key Findings

# Thank You
