---
title: "Modeling Body Fat"
subtitle: "L01G01"
author: "Lara Burns, Karin Gao, Qiuyu Liu, Bryan Chan and Hanna Boudara"
format: 
  revealjs:
    auto-slide: 20000
    embed-resources: true
    transition: slide
    css: "quarto presentation styles.css"
---

```{r, message=FALSE}
library(tidyverse)
library(readr)
# install.packages("modelsummary")
library(modelsummary)
library(ggcorrplot)
# install.packages("caTools")
library(caTools)
library(car)
# install.packages("sjPlot")
library(sjPlot)
library(insight)
# install.packages("lmSubsets")
library(lmSubsets)
library(broom)
theme_set(theme_bw())
```

## Motivation and Research Question

Research Question: How do body characteristics affect body fat percentage?

## Data Cleaning and Splitting

The original data had no missing values. For the variable `Weight`, which was in pounds, was converted to kilograms and `Height`, which was in inches, to centimeters. After this data cleaning, we had the following data set to build our models upon:

```{r}
# Read in data and inspect
bodyfat <- read_tsv("bodyfat.txt") 
```

```{r}
bodyfat$Age <- as.integer(bodyfat$Age)
bodyfat$Weight <- bodyfat$Weight/2.205
bodyfat$Height <- bodyfat$Height*2.54
glimpse(bodyfat)

```

Our models were built on 80% of the data set, designated as the training data, and the models' performance was built on the remaining 20% of the data, designated as the test data.

```{r}
set.seed(71) # Set random seed

split <- sample.split(bodyfat$Pct.BF, SplitRatio = 0.8)
train <- subset(bodyfat, split == TRUE)
test <- subset(bodyfat, split == FALSE)
```

## Exploratory Data Analysis

```{r}
# install.packages("patchwork")
library(patchwork)

# Histogram of Body Fat %
p1 <- ggplot(bodyfat, aes(x = Pct.BF)) +
  geom_histogram(binwidth = 3, boundary = 0,
                 fill = "lightblue", color = "white", alpha = 0.9) +
  labs(title = "Distribution of Body Fat Percentage",
       x = "Body Fat (%)", y = "Count") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5))

# Scatter plots
vars_to_plot <- c("Density", "Abdomen","Waist","Chest","Hip","Weight","Thigh",
                  "Neck","Knee","Bicep","Forearm","Wrist","Age","Ankle","Height")

p2 <- bodyfat %>%
  select(Pct.BF, all_of(vars_to_plot)) %>%
  pivot_longer(cols = all_of(vars_to_plot),
               names_to = "variable", values_to = "value") %>%
  drop_na() %>%
  ggplot(aes(x = value, y = Pct.BF)) +
  geom_point(alpha = 0.5, colour = "lightblue", size = 0.9) +
  geom_smooth(method = "lm", se = TRUE, colour = "red", linewidth = 0.7) +
  facet_wrap(~ variable, scales = "free_x", ncol = 5) +
  labs(x = NULL, y = "Body Fat (%)",
       title = "Relationships between Body Fat % and Predictors") +
  theme_classic() + 
  theme(legend.position = "top",
        strip.text = element_text(size = 10),
        plot.title = element_text(size = 14, hjust=0.5))

# Combining graphs together
p1 + p2 + plot_layout(widths = c(1, 2))
```

## Full Linear Model: $$\widehat{\text{Pct.BF}} = 4.22 + 0.07(\text{Age}) - 0.04(\text{Weight})- 0.10(\text{Height}) - 0.63(\text{Neck}) - 0.15(\text{Chest}) + 0.96(\text{Abdomen}) \\ - 0.24(\text{Hip}) + 0.30(\text{Thigh}) - 0.21(\text{Knee}) + 0.30(\text{Ankle}) + 0.14(\text{Bicep})+ 0.35(\text{Forearm}) -1.37(\text{Wrist})$$

```{r}
M1 = lm(Pct.BF ~.-Density, data=train) # Full Model without density
round(summary(M1)$coef, 3)
```

Our resulting model is:

$$\widehat{\text{Pct.BF}} = 484.598 + 0.008(\text{Age})+ 0.016(\text{Weight})- 0.021(\text{Height}) \\+ 0.027(\text{Neck})- 0.003(\text{Chest})- 0.006(\text{Abdomen})+ 0.024(\text{Hip})- 0.001(\text{Thigh}) \\- 0.048(\text{Knee})+ 0.016(\text{Ankle})- 0.003(\text{Bicep})+ 0.005(\text{Forearm})- 0.098(\text{Wrist})$$

```{r}
# Helper function
model_footer <- function(model) {
  s <- summary(model)
  
  # Extract p-value from F-statistic
  p_val <- pf(s$fstatistic[1], s$fstatistic[2], s$fstatistic[3], lower.tail = FALSE)
  
  # Print formatted summary
  cat(
    sprintf("Residual standard error: %.3f on %d degrees of freedom\n", s$sigma, s$df[2]),
    sprintf("Multiple R-squared: %.4f, Adjusted R-squared: %.4f\n", s$r.squared, s$adj.r.squared),
    sprintf("F-statistic: %.1f on %d and %d DF, p-value: %.3g\n",
            s$fstatistic["value"], s$fstatistic["numdf"], s$fstatistic["dendf"], p_val)
  )
}
```

```{r}
model_footer(M1)
```

## Full Linear Model (Continued)

In the full model, majority of predictors are statistically insignificant (with `Pr(>|t|)` \> 0.05) and this is potentially due to multi-collinearity between the predictors.

```{r}
alias(M1)
```

This shows that waist and abdomen are nearly perfectly correlated. This would make model predictions unstable, hence we can try different models selection methods.

## Model with Interaction Term

```{r}
lm_bh = lm(Pct.BF ~ Abdomen + Waist + Chest + Hip + Abdomen:Chest, data = train)
summary(lm_bh)
```

## Assumptions

```{r}
aug_bh = augment(lm_bh)

p1 <- ggplot(aug_bh, aes(.fitted, .std.resid)) +
geom_point(alpha = 0.6) +
geom_hline(yintercept = 0, color = "red") + 
labs(x = "Fitted values", y = "Standardized Residuals", title = "Residuals vs Fitted for Abdomen x Chest") +
geom_smooth(method = "loess", se = FALSE)

p2 <- ggplot(aug_bh, aes(sample = .std.resid)) +
stat_qq() +
stat_qq_line() +
labs(title = "Q-Q Plot of Residuals")
p1 + p2
```

```{r}
pred_bh = predict(lm_bh, newdata = test)
RMSE = function(actual, pred) sqrt(mean((actual - pred)^2))
rmse_bh = RMSE(test$Pct.BF, pred_bh)
rmse_bh

```

```{r}
lm_bh_noint = lm(Pct.BF ~ Abdomen + Waist + Chest + Hip, data = train)
anova(lm_bh_noint, lm_bh)  # Likelihood ratio test
AIC(lm_bh_noint, lm_bh)    # Compare the model selection criteria
```

```{r}
# Refit without interaction, bc high p-value
lm_bh_final <- lm(Pct.BF ~ Abdomen + Waist + Chest + Hip, data = train)

# RMSE
pred_bh_final <- predict(lm_bh_final, newdata = test)
rmse_bh_final <- RMSE(test$Pct.BF, pred_bh_final)
rmse_bh_final
```

## Multiple Body Sections Model: $\widehat{\text{Pct.BF}} = -18.10 + 0.85(\text{Abdomen}) + 0.02(\text{Bicep}) - 0.96(\text{Neck}) - 0.10(\text{Thigh})$

Examining correlations:

```{r}
correlations <- cor(bodyfat[, c("Pct.BF", "Abdomen", "Chest", "Waist", "Hip",
                                "Bicep", "Forearm", "Wrist",
                                "Neck", 
                                "Thigh", "Knee", "Ankle")], use = "complete.obs")

correlations["Pct.BF", ]
```

The variables most correlated with body fat percentage from each body section are: `Abdomen` (torso), `Bicep` (arms), `Neck` (neck/head), and `Thigh` (legs). The resulting model using these:

```{r}
lm1 <- lm(Pct.BF ~ Abdomen + Bicep + Neck + Thigh, data = train)
round(summary(lm1)$coef, 3)
```

Our resulting model is:

$\widehat{\text{Pct.BF}} = -18.10 + 0.85 \, \text{Abdomen} + 0.02 \, \text{Bicep} - 0.96 \, \text{Neck} - 0.10 \, \text{Thigh}$

## Backwards Selection: $$\widehat{\text{Pct.BF}} = 9.55 + 0.06(\text{Age}) - 0.114(\text{Height})- 0.63(\text{Neck}) -0.15(\text{Chest})+ 0.94(\text{Abdomen}) \\ - 0.27(\text{Hip}) + 0.29(\text{Thigh}) +0.37(\text{Forearm})- 1.22(\text{Wrist})$$

```{r}
step.back.aic = step(M1, 
					direction = "backward", 
					trace = FALSE) 
round(summary(step.back.aic)$coef, 3)
```

Our resulting model is:

$$\widehat{Pct.BF} = 9.551 - 0.114(Height) - 0.625(Neck) + 0.940(Abdomen) - 1.217(Wrist)$$

```{r}
model_footer(step.back.aic)
```

## Forward Selection: $$\widehat{\text{Pct.BF}} = -32.68  + 2.59(\text{ Waist}) - 0.28(\text{Weight}) - 0.98(\text{Wrist})\\ + 0.25(\text{Bicep}) - 0.51(\text{Neck}) + 0.33(\text{Forearm})$$

```{r}
M0 = lm(Pct.BF ~ 1, data=train) # Null model
step.fwd.aic = step(M0, scope = list(lower = M0, upper = M1),
                    direction = "forward", trace = FALSE)
round(summary(step.fwd.aic)$coef, 3)
```

Our resulting model is:

$$
\widehat{\text{Pct.BF}} = 486.473 - 439.635(\text{Density}) - 0.025(\text{Height}) + 0.014(\text{Weight})
$$

## Comparing Stepwise Models

```{r}
sjPlot::tab_model(step.fwd.aic, step.back.aic, 
                  show.ci = FALSE, 
                  show.aic = TRUE,
                  dv.labels = c("Forward model", "Backward model"))
```

## Exhaustive Search

```{r}
exh = lmSubsets(Pct.BF ~ ., data = train, nbest = 15)
plot(exh)
```

## Exhaustive Model: $$\widehat{\text{Pct.BF}} = 4.22 + 0.07(\text{Age}) - 0.04(\text{Weight})  - 0.10(\text{Height}) - 0.63(\text{Neck}) \\- 0.15(\text{Chest}) + 0.96(\text{Abdomen})  - 0.24(\text{Hip}) + 0.30(\text{Thigh}) - 0.21(\text{Knee}) \\+ 0.30(\text{Ankle}) +0.15(\text{Bicep})+ 0.35(\text{Forearm})- 1.37(\text{Wrist})$$

```{r}
best_exh <- lmSelect(exh, criterion = "AIC")
# Refit chosen model
best_model <- lm(update(formula(best_exh), . ~ . - Density), data = train)
round(summary(best_model)$coef, 3)
```

Our resulting model:

$$
\begin{aligned}\widehat{\text{Pct.BF}} = {} & 484.598- 438.310(\text{Density})+ 0.008(\text{Age})+ 0.016(\text{Weight})- 0.021(\text{Height}) \\& + 0.027(\text{Neck})- 0.003(\text{Chest})- 0.006(\text{Abdomen})+ 0.024(\text{Hip})- 0.001(\text{Thigh}) \\& - 0.048(\text{Knee})+ 0.016(\text{Ankle})- 0.003(\text{Bicep})+ 0.005(\text{Forearm})- 0.098(\text{Wrist})\end{aligned}
$$

```{r}
model_footer(best_model)
```

## Comparing All Models

```{r}
y_test <- test$Pct.BF

# Predictions on test set
pred_full <- predict(M1, newdata = test)
pred_fwd <- predict(step.fwd.aic, newdata = test)
pred_back <- predict(step.back.aic, newdata = test)
pred_exh <- predict(best_model, newdata = test)
pred_mbp <- predict(lm1, newdata = test)

# Helper function
RMSE <- function(actual, predicted) sqrt(mean((actual - predicted)^2))
MAE <- function(actual, predicted) mean(abs(actual - predicted))
R2 <- function(actual, predicted) cor(actual, predicted)^2

results <- data.frame(
Model = c("Full Model", "Multiple Body Parts", "Forward Stepwise",
"Backward Stepwise", "Exhaustive (BIC)"),
RMSE = c(RMSE(y_test, pred_full),
RMSE(y_test, pred_mbp),
RMSE(y_test, pred_fwd),
RMSE(y_test, pred_back),
RMSE(y_test, pred_exh)),
MAE = c(MAE(y_test, pred_full),
MAE(y_test, pred_mbp),
MAE(y_test, pred_fwd),
MAE(y_test, pred_back),
MAE(y_test, pred_exh)),
R2 = c(R2(y_test, pred_full),
R2(y_test, pred_mbp),
R2(y_test, pred_fwd),
R2(y_test, pred_back),
R2(y_test, pred_exh))
)
results <- results[order(results$RMSE), ]
rownames(results) <- NULL
print(results)
```

## Best Overall Model

-   List the best model's equation

-   Interpret coefficient signs and magnitudes

-   Discuss statistically significant predictors

## Assumptions of the Model

## Graphing of Residuals + QQ Plot (To be used for best overall model)

```{r}
# install.packages("broom")
library(broom)

aug <- augment(lm1)

p1 <- ggplot(aug, aes(.fitted, .std.resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "blue") +
  labs(x = "Fitted", y = "Std. Residual")

p2 <- ggplot(aug, aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line()

# Combine plots side by side
p1 + p2
```

## Key Findings

-   Discuss interesting things about the best overall model

-   Relate to research question

## Limitations

-   Include recommendations for future research

# Thank You
