---
title: "Modeling Body Fat"
subtitle: "How Do Body Characteristics Affect Body Fat Percentage?"
author: "L01G01: Lara Burns, Karin Gao, Qiuyu Liu, Bryan Chan and Hanna Boudara"
format: 
  revealjs:
    auto-slide: 20000
    embed-resources: true
    transition: slide
    css: "quarto presentation styles.css"
---

```{r, message=FALSE}
library(tidyverse)
library(readr)
# install.packages("modelsummary")
library(modelsummary)
library(ggcorrplot)
# install.packages("caTools")
library(caTools)
library(car)
# install.packages("sjPlot")
library(sjPlot)
library(insight)
# install.packages("lmSubsets")
library(lmSubsets)
theme_set(theme_bw())
```

## Data Cleaning and Splitting

The original data had no missing values. For the variable `Weight`, which was in pounds, was converted to kilograms and `Height`, which was in inches, to centimeters. After this data cleaning, we had the following data set to build our models upon:

```{r}
# Read in data and inspect
bodyfat <- read_tsv("bodyfat.txt") 
```

```{r}
bodyfat$Age <- as.integer(bodyfat$Age)
bodyfat$Weight <- bodyfat$Weight/2.205
bodyfat$Height <- bodyfat$Height*2.54
glimpse(bodyfat)

```

Our models were built on 80% of the data set, designated as the training data, and the models' performance was built on the remaining 20% of the data, designated as the test data.

```{r}
set.seed(71) # Set random seed

split <- sample.split(bodyfat$Pct.BF, SplitRatio = 0.8)
train <- subset(bodyfat, split == TRUE)
test <- subset(bodyfat, split == FALSE)
```

## Exploratory Analysis of the Data

```{r}
# install.packages("patchwork")
library(patchwork)  

# Histogram of Body Fat %
p1 <- ggplot(bodyfat, aes(x = Pct.BF)) +
  geom_histogram(binwidth = 3, boundary = 0,
                 fill = "lightblue", color = "white", alpha = 0.9) +
  labs(title = "Distribution of Body Fat Percentage",
       x = "Body Fat (%)", y = "Count") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5))

# Scatter plots
vars_to_plot <- c("Density", "Abdomen","Waist","Chest","Hip","Weight","Thigh",
                  "Neck","Knee","Bicep","Forearm","Wrist","Age","Ankle","Height")

p2 <- bodyfat %>%
  select(Pct.BF, all_of(vars_to_plot)) %>%
  pivot_longer(cols = all_of(vars_to_plot),
               names_to = "variable", values_to = "value") %>%
  drop_na() %>%
  ggplot(aes(x = value, y = Pct.BF)) +
  geom_point(alpha = 0.5, colour = "lightblue", size = 0.9) +
  geom_smooth(method = "lm", se = TRUE, colour = "red", linewidth = 0.7) +
  facet_wrap(~ variable, scales = "free_x", ncol = 5) +
  labs(x = NULL, y = "Body Fat (%)",
       title = "Relationships between Body Fat % and Predictors") +
  theme_classic() + 
  theme(legend.position = "top",
        strip.text = element_text(size = 10),
        plot.title = element_text(size = 14, hjust=0.5))

# Combining graphs together
p1 + p2 + plot_layout(widths = c(1, 2))
```

## Multiple Body Sections: Examining Correlations

```{r}
lm1 <- lm(Pct.BF ~ Abdomen + Bicep + Neck + Thigh, data = train)
lm1
```

Generated regression statistics:

```{r}
summary(lm1)
```

## Multiple Body Sections: Graphing of Residuals + QQ Plot

```{r}
# install.packages("broom")
library(broom)
# install.packages("patchwork")
library(patchwork)

aug <- augment(lm1)

p1 <- ggplot(aug, aes(.fitted, .std.resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "blue") +
  labs(x = "Fitted", y = "Std. Residual")

p2 <- ggplot(aug, aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line()

# Combine plots side by side
p1 + p2
```
